# Reproduceable research project 1
## Jim Smith
========================================================

The activity.csf file was extracted from repdata-data-activity.zip and placed in the working directory.
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
```{r input}
activityData = read.csv("activity.csv")
# use date format
activityData$date = as.Date(activityData$date, format = "%Y-%m-%d")
# use factors for interval
activityData$interval <- factor(activityData$interval)
```

Display a histogram of the steps taken each day along with mean and median steps

```{r firstHistogram}
# use aggregate to get a data frame of sum of steps by date
stepsByDay <- aggregate(steps ~ date, data=activityData, FUN=sum, na.rm=TRUE)
hist(stepsByDay$steps)
meanSteps = mean(stepsByDay$steps, na.rm=TRUE)
summary(stepsByDay$steps)

```
Plot of the average daily activity pattern?

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r plotIntervalAverages}
#install.packages("ggplot2")           #ggplot2 has to be installed once
library(ggplot2)
avgStepsByInterval <- aggregate(steps ~ interval, data=activityData, FUN=mean, na.rm=TRUE)
maxAvgStepsIndex <- which.max(avgStepsByInterval$steps)
avgStepsByInterval$steps[maxAvgStepsIndex]
avgStepsByInterval$interval[maxAvgStepsIndex]
# need to convert the factor back to numeric
qplot(as.numeric(as.character(interval)), steps, data=avgStepsByInterval, geom = 'line')
```

Count the  missing values, then compute new summary statistics and plot histogram

```{r replaceNAs}
#number of intervals that have NA
length(activityData$steps[is.na(activityData$steps)])
# the is.na creates a true/false vector, true for each NA. 
# $interval returns one of the 288 factor variables at each place their is an NA
# $steps returns an average number of steps based on the factor variable
# finally this is saved back into the original data table
activityData$steps[is.na(activityData$steps)] <- avgStepsByInterval$steps[activityData$interval[is.na(activityData$steps)]]
# replot and calculate means after replacing NAs with averages
stepsByDay <- aggregate(steps ~ date, data=activityData, FUN=sum, na.rm=TRUE)
hist(stepsByDay$steps)
meanSteps = mean(stepsByDay$steps, na.rm=TRUE)
summary(stepsByDay$steps)
```
The above statistics show that the histogram doesn't change very much. The mean and median
shift down a small amount from 10880 to 10770.

Are there differences in activity patterns between weekdays and weekends?

```{r weekVsWeekend}
weekends = c("Saturday" , "Sunday")
# adds a column with two factors, weekday and weekend
activityData <- cbind(activityData, ifelse(weekdays(activityData$date) %in% weekends, "weekend", "weekday"))
colnames(activityData)[4] <- "weekdays"
avgWeekdaysInterval <- aggregate(steps ~ interval + weekdays, data = activityData, mean)
qplot(as.numeric(as.character(interval)) , steps, data = avgWeekdaysInterval, 
      facets = weekdays ~ . , geom = 'line', xlab = "interval", ylab = "number of steps")
```
